---
title: "Partial Dependence Explanations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Dependence Explanations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the Partial Dependence explanations in `survex`, as well as Accumulated Local Effects explanations. It especially demonstrates the usage of new kinds of plots that are available in the 1.1 version of the package. To create these explanations we follow the standard way of working with survex i.e. we create a model, and an explainer.

```{r setup}
library(survex)
library(survival)
library(ranger)

set.seed(123)

vet <- survival::veteran

rsf <- ranger(Surv(time, status) ~ ., data = vet)
exp <- explain(rsf, data = vet[, -c(3,4)], y = Surv(vet$time, vet$status))

```

We use the explainer and the `model_profile()` function to calculate Partial Dependence explanations. We can specify the variables for which we want to calculate the explanations. In this example we calculate the explanations for the variables `karno` and `celltype`. **Note:** The background for generating PD values is the `data` field of the explainer! If you want to calculate explanations with a background that is not the training data, you need to manually specify the `data` argument, when creating the explainer.

We can calculate Accumulated Local Effects in the same way, by setting the `type` argument to `"accumulated"`.

```{r}
pdp <- model_profile(exp, variables = c("karno", "celltype"), n = 20)
ale <- model_profile(exp, variables = c("karno", "celltype"), n = 20, type = "accumulated")
```

To plot these explanations you can use the plot function. By default the explanations for all calculated variables are plotted. This example demonstrates this for the `pdp` object which contains the explanations for 2 variables.

```{r}
plot(pdp)
```

We can plot ALE explanations in the same way as PD explanations, as is demonstrated by the example below. For the rest of the vignette we only focus on PD explanations.

```{r}
plot(ale)
```

The `plot()` function can also be used to plot the explanations for a subset of variables. The `variables` argument specifies the variables for which the explanations are plotted. The `numerical_plot_type` argument specifies the type of plot for numerical variables. For `numerical_plot_type = "lines"` (default), the y-axis represents the mean prediction (survival function), x-axis represents the time dimension, and different colors represent values of the studied variable. For `numerical_plot_type = "contours"`, the y-axis represents the values of the studied variable, x-axis represents time, and different colors represent the mean prediction (survival function).

```{r}
plot(pdp, variables = c("karno"), numerical_plot_type = "contours")
```

The plots above make use of the time dependent output of survival models, by placing the time dimension on the x-axis. However, for people familiar with Partial Dependence explanations in classification and regression, it might be more intuitive to place the variable values on the x-axis. For this reason, we provide the `plot2()` function, which can display the explanations without the aspect of time. 

To use this function a specific time of interest has to be chosen. This time needs to be one of the values in the `times` field of the explainer. If the automatically generated times do not contain the time of interest, one needs to manually specify the `times` argument when creating the explainer.

The example below shows the PD explanations for the `karno` variable at time 80. The y-axis represents the mean prediction (survival function), x-axis represents the values of the studied variable. Thin background lines are individual ceteris paribus profiles (otherwise known as ICE profiles).

```{r}
plot2(pdp, variable = "karno", times = 80)
```

The same plot can be generated for the categorical `celltype` variable. In this case the x-axis represents the different values of the studied variable, boxplots present the distribution of individual ceteris paribus profiles, and the line represents the mean prediction (survival function), which is the PD explanation.

```{r}
plot2(pdp, variable = "celltype", times = 80)
```

Of course, the plots can be prepared for multiple timepoints, at the same time and presented on one plot.

```{r}
plot2(pdp, variable = "karno", times = c(1, 80, 151.72))
```

```{r}
plot2(pdp, variable = "celltype", times = c(1, 80, 151.72))
```


`survex` also implements 2 dimensional PD and ALE explanations. These can be used to study the interaction between variables. The `model_profile_2d()` function can be used to calculate these explanations. The `variables` argument specifies the variables for which the explanations are calculated. The `type` argument specifies the type of explanation, and can be set to `"partial"` (default) or `"accumulated"`.

```{r}
pdp_2d  <- model_profile_2d(exp, variables = list(c("karno", "age")))
pdp_2d_num_cat <- model_profile_2d(exp, variables = list(c("karno", "celltype")))
```

These explanations can be plotted using the plot function.  

```{r}
plot(pdp_2d, times = 80)
```

```{r}
plot(pdp_2d_num_cat, times = 80)
```

