---
title: "How to use survex with mlr3proba"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### mlr3proba and additional packages 

The `mlr3proba` package maintainers state that 

> mlr3proba is not on CRAN and is unlikely to be reuploaded (see [here](https://twitter.com/RaphaelS101/status/1506321623250571265) for reasons). 

For this reason, the installation of this package may be not obvious. The following chunk loads all required packages for this vignette.

```{r setup, message=FALSE, results=FALSE}

if (!require("ooplah")) install.packages("ooplah",
                                         repos = "https://mlr-org.r-universe.dev")

if (!require("param6")) install.packages("param6",
                                         repos = "https://mlr-org.r-universe.dev")

if (!require("set6")) install.packages("set6",
                                         repos = "https://mlr-org.r-universe.dev")

if (!require("distr6")) install.packages("distr6",
                                         repos = "https://mlr-org.r-universe.dev")

if (!require("mlr3")) install.packages("mlr3")

if (!require("mlr3proba")) install.packages("mlr3proba",
                                            repos = "https://mlr-org.r-universe.dev")
if (!require("mlr3extralearners")) install.packages("mlr3extralearners",
                                                    repos = "https://mlr-org.r-universe.dev")
if (!require("mlr3pipelines")) install.packages("mlr3pipelines")

library(mlr3proba)
library(mlr3extralearners)
library(mlr3pipelines)
library(survex)
library(survival)
```

### Creating an explainer for mlr3proba learner

`survex` allows for the creation of explainers for `mlr3proba::LearnerSurv()` objects in the following way:

```{r}
veteran_task <- as_task_surv(veteran,
                     time = "time",
                     event = "status",
                     type = "right")
ranger_learner <- lrn("surv.ranger")	
ranger_learner$train(veteran_task)
ranger_learner_explainer <- explain(ranger_learner, 
                     data = veteran[, -c(3,4)],
                     y = Surv(veteran$time, veteran$status),
                     label = "Ranger model")
```

#### Example explanation 

The created explainer can be easily used for providing explanations.

```{r}
ranger_learner_explainer |> predict_profile(veteran[1,]) |> plot(numerical_plot_type = "contours",
                                                  variables = c("karno", "celltype"),
                                                  facet_ncol = 1,
                                                  subtitle = NULL)
```

### Creating explainer for learner with composed distribution prediction

Not all models can predict the survival distribution (`distr` predict type), however they can be combined with baseline survival distribution estimator to create composite models. Such models created using the `distrcompositor()` pipeline can also be explained easily. There is one caveat, as the pipeline strips the `LearnerSurv()` class, it has to be added again manually.

``` {r message=FALSE}
gbm_composite_learner <- as_learner(ppl(
    "distrcompositor",
    learner = lrn("surv.gbm"),
    estimator = "kaplan",
    form = "ph"
))
gbm_composite_learner$train(veteran_task)

# important!
class(gbm_composite_learner) <- c(class(gbm_composite_learner), "LearnerSurv") 
gbm_composite_learner_explainer <- explain(gbm_composite_learner,
                                       data = veteran[, -c(3,4)],
                                       y = Surv(veteran$time, veteran$status),
                                       label = "Composite GBM model")

gbm_composite_learner_explainer |> model_parts(type = "ratio") |> plot(subtitle = NULL)
```

### Using mlr3proba measures

The `mlr3proba` package provides many additional performance measures. They can be adapted for use with `survex` using the `loss_adapt_mlr3proba()` function. These adapted metrics can be used with `model_parts()` and `model_performance()` functions.
``` {r message=FALSE}
loss_schmid <- loss_adapt_mlr3proba(msr("surv.schmid"))
gbm_composite_learner_explainer |> model_parts(loss = loss_schmid) -> model_part
model_part |> plot(subtitle = NULL)
```

``` {r message=FALSE}
loss_rcll <- loss_adapt_mlr3proba(msr("surv.rcll"))
custom_metrics <- c("Integrated Schmid Score" = loss_schmid, 
                    "Right-Censored Log loss" = loss_rcll)

perf_ranger <- model_performance(ranger_learner_explainer, metrics=custom_metrics)
perf_comp_gbm <- model_performance(gbm_composite_learner_explainer, metrics=custom_metrics) 

plot(perf_ranger, perf_comp_gbm, metrics_type="scalar", subtitle = NULL)
```

